---
title: "ENVS-193DS_homework-03"
format: html
name: "Linda Herrera"
date: "2026-02-25"
---

## Part 1. Set up tasks 

```{r}
#| label: libraries and data
#| message: false
library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # generating model predictions
library(gtsummary) # generating summary tables for models

# read in data as object "pickleweed"
pickleweed <- read.csv(here("data", "salinity-pickleweed.csv"))
# read in personal data as object 
my_data <- read_xlsx(here("data", "ENVS193_personal_data.xlsx"))
```

## Part 2. Problems

### Problem 1. Slough soil salinity

#### a. An appropriate test

#### b. Create a visualization

```{r}
#| label: ggplot
#| message: false


```

#### c. Check your assumptions and run your test

```{r}
#| label: test
#| message: false

```

#### d. Results communication

#### e. Test implications

#### f. Double check your own work.

```{r}
#| label: test
#| message: false


```


### Problem 2. Personal 

#### a. Updating your visualizations

```{r}
#| label: cleaning and wrangling 

```

```{r}
#| label: ggplot 1
#| message: false

```

```{r}
#| label: ggplot 2
#| message: false

```

#### b. Captions


### Problem 3. Affective visualization

#### a. Describe in words what an affective visualization could look like for your personal data (3-5 sentences). 

#### b. Create a sketch (on paper) of your idea.

#### c. Make a draft of your visualization.

#### d. Write an artist statement.

#### e. Prep your materials to share in class.


### Problem 4. Statistical critique

#### a. Revisit and summarize

#### b. Visual clarity

#### c. Aesthetic clarity

#### d. Recommendations


