---
title: "ENVS-193DS_homework-03"
format: html
name: "Linda Herrera"
date: "2026-02-25"
---

## Part 1. Set up tasks 

```{r}
#| label: libraries and data
#| message: false
library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # generating model predictions
library(gtsummary) # generating summary tables for models
library(ggplot2) # making ggplots 
library(viridisLite)

# read in data as object "pickleweed"
pickleweed <- read.csv(here("data", "salinity-pickleweed.csv"))
# read in personal data as object 
my_data <- read_xlsx(here("data", "ENVS193_personal_data.xlsx"))
```

## Part 2. Problems

### Problem 1. Slough soil salinity

#### a. An appropriate test

The two appropriate tests are Pearson's correlation coefficient (r) and Spearman's rank correlation coefficient (ρ), which both quantify the strength and direction of the relationship between two continuous variables: salinity (e.g., soil or water salinity concentration) and California pickleweed biomass (e.g., grams of dry mass). Pearson’s r tests for a linear relationship and assumes approximately normally distributed, continuous variables with independent observations, whereas Spearman’s ρ is a non-parametric, rank-based test that assesses a monotonic relationship without requiring normality. 

#### b. Create a visualization

```{r}
#| label: ggplot
#| message: false


```

#### c. Check your assumptions and run your test

```{r}
#| label: test
#| message: false

```

#### d. Results communication

#### e. Test implications

#### f. Double check your own work.

```{r}
#| label: test
#| message: false


```


### Problem 2. Personal 

#### a. Updating your visualizations

```{r}
#| label: cleaning and wrangling
#| message: false

my_data_clean <- na.omit(my_data) |> # remove all the NA's within personal data 
clean_names() |> # clean the data names
  rename(date = "date_yyyy_mm_dd", # renaming all the columns to be short and simple 
         dow = "day_of_week",
         hw = "class_homework_time_hrs", 
         bible_study = "bible_study_completed_yes_no",
         min = "minutes_of_bible_study_min",
         location = "location_home_ucen_church_building_other" ,
         time = "time_of_day", 
         with_someone = "with_someone_yes_no") |> 
  mutate(hw_mins = hw*60)

```

```{r}
#| label: ggplot 1
#| message: false

ggplot(data = my_data_clean, # start with the my_data_clean data frame
       aes(x = with_someone, # categorical predictor variable on the x-axis
           y = min, # response variable on the y-axis
           color = with_someone)) + # coloring by production type
  geom_boxplot() + # first layer should be a boxplot
  geom_jitter(width = 0.2, # making the points jitter horizontally
              height = 0) + # making sure points don't jitter vertically
  labs(x = "With Someone (Yes/No)", # label x- and y- axes and add a title
       y = "Bible Study Devotional (mins)",
       title = "Bible Time When Studying Alone vs. With Others") +
   scale_color_manual(values = c("No" = "red", # custom change the colors
                                "Yes" = "green4")) +
  theme_light() + # change the theme 
  theme(legend.position = "none") # remove the legend 

```

```{r}
#| label: ggplot 2
#| message: false

ggplot(data = my_data_clean, # start with the my_data_clean data frame
       aes(x = hw_mins, # categorical predictor variable on the x-axis
           y = min,
           shape = dow)) + # response variable on the y-axis
  geom_point(color = "purple", # add points and change the color 
             size = 2) + 
  labs(x = "Time spent doing Homework/Classes (mins)", # label x- and y- axes and add a title 
       y = "Bible Study Devotional (mins)",
       title = "Academic Time vs. Bible Study") + 
  theme_light() # change the theme 

```

```{r}
my_data_clean |> 
  arrange(desc(min)) |> 
  ggplot(aes(x = hw_mins,
             y = min,
             size = min,
             fill = location)) +
  geom_point(alpha = 0.5,
             shape = 21,
             color = "black") +
  scale_size(range = c(1, 15), name = "Bible Study (mins)") +
  scale_fill_viridis_d(option = "A", guide = "none") +
  theme_light() +
  labs(x = "Time spent doing Homework/Classes (mins)",
       y = "Bible Study Devotional (mins)",
       title = "Academic Time vs. Bible Study") +
  theme(legend.position = "bottom")

```



#### b. Captions


### Problem 3. Affective visualization

#### a. Describe in words what an affective visualization could look like for your personal data (3-5 sentences). 

#### b. Create a sketch (on paper) of your idea.

#### c. Make a draft of your visualization.

#### d. Write an artist statement.

#### e. Prep your materials to share in class.


### Problem 4. Statistical critique

#### a. Revisit and summarize

#### b. Visual clarity

#### c. Aesthetic clarity

#### d. Recommendations


