---
title: "ENVS-193DS_homework-03"
format: html
name: "Linda Herrera"
date: "2026-02-25"
---

## Part 1. Set up tasks 

```{r}
#| label: libraries and data
#| message: false
library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # generating model predictions
library(gtsummary) # generating summary tables for models
library(ggplot2) # making ggplots 
library(viridisLite)
library(flextable)

# read in data as object "pickleweed"
pickleweed <- read.csv(here("data", "salinity-pickleweed.csv"))
# read in personal data as object 
my_data <- read_xlsx(here("data", "ENVS193_personal_data.xlsx"))
```

## Part 2. Problems

### Problem 1. Slough soil salinity

#### a. An appropriate test

The two appropriate tests are Pearson's correlation coefficient (r) and Spearman's rank correlation coefficient (ρ), which both quantify the strength and direction of the relationship between two continuous variables: salinity (e.g., soil or water salinity concentration) and California pickleweed biomass (e.g., grams of dry mass). Pearson’s r tests for a linear relationship and assumes approximately normally distributed, continuous variables with independent observations, whereas Spearman’s ρ is a non-parametric, rank-based test that assesses a monotonic relationship without requiring normality. 

#### b. Create a visualization

```{r}
#| label: cleaning
#| message: false
pickleweed_clean <- pickleweed |> 
  clean_names() |> 
  select(salinity_m_s_cm, pickleweed) |> 
  rename(salinity = salinity_m_s_cm,
         biomass = pickleweed)

```

```{r}
#| label: pickleweed scatterplot
#| message: false
#| fig-height: 4
#| fig-width: 6
ggplot(data = pickleweed_clean, 
       aes(x = salinity,
           y = biomass)) + 
  geom_point(size = 3,
             color = "seagreen3",
             alpha = 21) +
  labs(x = "Soil Salinity (mS/cm)",
       y = "California Pickleweed Biomass (g)") +
  theme_minimal()

```

#### c. Check your assumptions and run your test

##### 1) Check assumptions:
```{r}
#| label: pickleweed-model
#| message: false
pickleweed_lm <- lm(
  biomass ~ salinity, 
  data = pickleweed_clean)

```

```{r}
#| label: pickleweed-model-diagnostics
#| message: false
par(mfrow = c(2, 2))
plot(pickleweed_lm) 

```

```{r}
#| label: model estimates in table
#| message: false

as_flextable(pickleweed_lm) 
```
Table 1. Results of simple linear regression examining the relationship between soil salinity (mS/cm) and California pickleweed biomass (g). 

```{r}
#| label: pickleweed-model-predictions
#| message: false

pickleweed_preds <- ggpredict(
  pickleweed_lm, # model object
  terms = "salinity") # predictor (in quotation marks)

pickleweed_preds
```

```{r}
#| label: pickleweed-model-final-figure
#| message: false

ggplot(data = pickleweed_clean,
       aes(x = salinity,
           y = biomass)) +
  geom_point(size = 3,
             stroke = 0.2,
             fill = "seagreen3",
             shape = 21) +
  geom_ribbon(data = pickleweed_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.1) +
  geom_line(data = pickleweed_preds,
            aes(x = x,
                y = predicted)) +
  labs(x = "Soil Salinity (mS/cm)", 
       y = expression("California Pickleweed Biomass (g)")) +
  theme_minimal()

```

To determine whether Pearson’s correlation was appropriate for testing the relationship between soil salinity (mS/cm) and California pickleweed biomass (g), I checked the assumptions of linearity, normality of residuals, homoscedasticity, and influential outliers. I evaluated these using the scatterplot and linear model diagnostic plots, including the Residuals vs. Fitted plot for linearity and constant variance, the Normal Q–Q plot for normality, and the Residuals vs. Leverage plot with Cook’s distance for influential points. The diagnostics indicated a linear relationship, reasonably normal residuals, constant variance, and no influential outliers, so the assumptions for Pearson’s correlation were met.

##### 2) Run test:
```{r}
#| label: pearson test
#| message: false
cor.test(pickleweed_clean$salinity, pickleweed_clean$biomass,
         method = "pearson")

```

#### d. Results communication

To evaluate the strength of the relationship between California pickleweed biomass (g) and soil salinity (mS/cm), I used Pearson’s correlation, which is appropriate for testing the linear relationship between two continuous variables. We found a moderate positive relationship between soil salinity and pickleweed biomass (Pearson’s r = 0.53, t(21) = 2.90, p = 0.0086; R² = 0.29; Table 1), indicating that soil salinity significantly predicts biomass.

From our linear model coefficient estimates, we found that for each 1 mS/cm increase in soil salinity, pickleweed biomass increases by approximately 2.08 ± 0.72 g (Table 1). Biologically, this suggests that larger pickleweed individuals are more likely to occur in higher-salinity soils, while smaller individuals are more likely to be found in lower-salinity soils.

#### e. Test implications

You’re working on a team of people at this restoration site who are also concerned about pickleweed planting. In 2-3 sentences, write what you would communicate to them about the results of this test and what it means for pickleweed planting success at your site.

Be cognizant of your audience as you are writing: what would they need to know to take action?

The results of this test show that pickleweed grows best in areas with higher salinity, confirming that it is well adapted to salty marsh conditions. This means we should prioritize planting pickleweed in zones of our restoration site that regularly receive tidal input and maintain moderate to high salinity levels. Focusing efforts in these areas will likely improve survival and overall planting success.

#### f. Double check your own work.

```{r}
#| label: test
#| message: false
cor.test(pickleweed_clean$salinity, pickleweed_clean$biomass,
         method = "spearman")

```

Both tests would lead to the same decision about the null hypothesis because both show statistically significant associations between soil salinity and pickleweed biomass (Spearman: ρ = 0.593, p = 0.0034; Pearson: r = 0.67, p = 0.0046). While Spearman measures a monotonic relationship using ranked salinity and biomass values and Pearson measures a linear relationship using the raw continuous variables, both indicate a statistically significant, moderate positive relationship, meaning pickleweed biomass increases as soil salinity increases.

### Problem 2. Personal 

#### a. Updating your visualizations

```{r}
#| label: cleaning and wrangling
#| message: false

my_data_clean <- na.omit(my_data) |> # remove all the NA's within personal data 
clean_names() |> # clean the data names
  rename(date = "date_yyyy_mm_dd", # renaming all the columns to be short and simple 
         dow = "day_of_week",
         hw = "class_homework_time_hrs", 
         bible_study = "bible_study_completed_yes_no",
         min = "minutes_of_bible_study_min",
         location = "location_home_ucen_church_building_other" ,
         time = "time_of_day", 
         with_someone = "with_someone_yes_no") |> 
  mutate(hw_mins = hw*60)

```

```{r}
#| label: ggplot 1
#| message: false

ggplot(data = my_data_clean, # start with the my_data_clean data frame
       aes(x = with_someone, # categorical predictor variable on the x-axis
           y = min, # response variable on the y-axis
           fill = with_someone)) + # coloring by production type
  geom_boxplot(alpha = 0.7, # first layer should be a boxplot
               width = 0.5,
               outlier.shape = NA) + 
  geom_jitter(width = 0.2, # making the points jitter horizontally
              height = 0,
              alpha = 0.5,
              size = 2,
              color = "black") + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 23,
               size = 3,
               fill = "white") +
  scale_fill_manual(values = c("No" = "orangered3",
                               "Yes" = "limegreen")) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0, .05))) +
  labs(x = "With Someone (Yes/No)", # label x- and y- axes and add a title
       y = "Bible Study Devotional (mins)",
       title = "Bible Time When Studying Alone vs. With Others",
       subtitle = "Most recent observation: 25 Feb 2026") +
  theme_minimal(base_size = 12) + # change the theme 
  theme(legend.position = "none",
         panel.grid.minor = element_blank(),
         plot.title = element_text(face = "bold")) # remove the legend 

```


```{r}
#| label: ggplot 2
#| message: false
#| fig-height: 4
#| fig-width: 6

ggplot(my_data_clean,
       aes(x = hw_mins,
           y = min,
           shape = location,
           fill = location)) +
  geom_point(size = 4,
             alpha = 0.9,
             color = "black") + 
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) + 
  scale_fill_viridis_d(option = "G") +
  coord_cartesian(ylim = c(0, 130)) +
  labs(
    x = "Time spent doing Homework/Classes (mins)",
    y = "Bible Study Devotional (mins)",
    title = "Daily Homework Time vs. Bible Study",
    subtitle = "Most recent observation: 25 Feb 2026") +
  theme_light() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"))

```

#### b. Captions


### Problem 3. Affective visualization

#### a. Describe in words what an affective visualization could look like for your personal data (3-5 sentences). 

#### b. Create a sketch (on paper) of your idea.

#### c. Make a draft of your visualization.

#### d. Write an artist statement.

#### e. Prep your materials to share in class.


### Problem 4. Statistical critique

#### a. Revisit and summarize

#### b. Visual clarity

#### c. Aesthetic clarity

#### d. Recommendations


